# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data


```{r, echo = T}
# Set working directory
setwd("/Users/emelaktas/RepData_PeerAssessment1")
```




```{r, echo = T}
# From assignment instructions: Load the data (i.e. read.csv())

# Load the data using read.table()
data <- read.table(file = "activity.csv", sep = ",", header = T)

# Summary of data
summary(data)

# Check the sstructure of variables
str(data)

# From assignment instructions: Process/transform the data (if necessary) into a format suitable for your analysis

# Convert date to date (it is read as a factor) datev: date variable, not to override the original date.
data$datev <- as.Date(data$date, format = "%Y-%m-%d")

# Check the sstructure of variables again and see datev is date
str(data)

# Summary of data
summary(data)

```


## What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the dataset.

```{r, echo = T}

# Since we can ignore missing values, let us work with complete cases in the data set.
datacomplete <-data[complete.cases(data),]

# Create the sum of steps for each day. First argument in tapply() is the variable sum of which is to be calculated. Second argument in tapply() is the date variable, since we sum the steps for each day. Third argument in tapply() is the function we want to apply on the first variable. 
totalsteps <- tapply(datacomplete$steps, datacomplete$datev, sum)

# From assignment instructions: Make a histogram of the total number of steps taken each day
hist(totalsteps, main = "Histogram of Total Number of Steps", xlab = "")

# From assignment instructions: Calculate and report the mean and median total number of steps taken per day

# Calculate mean
meansteps <- mean(totalsteps)
meansteps

# Calculate median
mediansteps <- median(totalsteps)
mediansteps
```

The mean value of the total number of steps per day is `r sprintf("%.0f",meansteps)`. The median of total number of steps taken per day is `r mediansteps`.

## What is the average daily activity pattern?

```{r, echo = T}
# Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

# Use tapply to calculate average number of steps in the interval averaged across all days
averagesteps <- tapply(datacomplete$steps, datacomplete$interval, mean)

# Plot time series: x = interval, y = average number of steps
plot(names(averagesteps), averagesteps, type = "l", xlab="", main = "Average number of steps per 5-minute interval", ylab = "Average steps")

# Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
interval.max <- averagesteps[which(averagesteps == max(averagesteps))]
interval.max

# Lower limit of the interval
interval.lower <- names(averagesteps)[which(averagesteps == max(averagesteps))-1]
interval.lower

# Upper limit of the interval
interval.upper <- names(averagesteps)[which(averagesteps == max(averagesteps))]
interval.upper
```


The interval `r interval.lower` - `r interval.upper` contains the maximum number of steps on average across all the days in the dataset. The maximum of average number of steps is `r interval.max`.


## Imputing missing values

```{r, echo = T} 
# From assignment instructions: Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
total.missing <- sum(is.na(data$steps))
total.missing

# From assignment instructions:  Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

# Replacing with the mean for the same interval may make more sense for week days. 
# Create a variable that shows the day of the week
data$day <-  weekdays(data$datev)

# Update datacomplete 
datacomplete <-data[complete.cases(data),]

# Use reshape to melt data
require(reshape)

# Melt data
data.molten <- melt.data.frame(datacomplete, id.vars = c("datev", "interval", "day"), measure.vars = "steps", variable_name ="steps")

# Cast data, calculates mean of steps for each interval * day combination
imputed.mean <- cast(data.molten, interval + day ~ steps, mean)

# Create a new dataset that is equal to the original dataset but with the missing data filled in.
data.imputed <- data.frame(steps = numeric(dim(data)[1]), interval = data$interval, date = data$datev, day = data$day)

# Replace NAs with calculated mean of interval by day
for (i in 1:dim(data)[1]){
  if(is.na(data$steps[i])){
    data.imputed$steps[i] <- imputed.mean$steps[which(imputed.mean$interval == data$interval[i] & imputed.mean$day == data$day[i])]
  }
  }

# From assignment instructions: Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

# Total number of steps with NAs replaced with mean steps for the day and interval
totalsteps.imputed <- tapply(data.imputed$steps, data.imputed$date, sum)

# Histogram of total number of steps 
hist(totalsteps.imputed, main = "Histogram of Total Number of Steps NAs imputed", xlab = "")

# Mean total number of steps
meansteps.imputed <- mean(totalsteps.imputed)
meansteps
meansteps.imputed

if(meansteps == meansteps.imputed) {
  print("Mean of total steps with NAs imputed is equal to the mean of total steps calculated with NAs removed.")} else {
    print("Mean of total steps has changed after imputing mean steps for the day and interval for NAs.")}

mediansteps.imputed <- median(totalsteps.imputed)
mediansteps
mediansteps.imputed

if(mediansteps == mediansteps.imputed) {
  print("Median of total steps with NAs imputed is equal to the mean of total steps calculated with NAs removed.")} else {
    print("Median of total steps has changed after imputing mean steps for the day and interval for NAs.")}

```

The total number of missing values in the dataset is `r total.missing`. When NAs are imputed with mean of the corresponding day and interval, they decrease the mean of observed variables because the number of NAs are large:  `r round(total.missing / dim(data.imputed)[1] *100,2)` per cent of the data is missing.

## Are there differences in activity patterns between weekdays and weekends?

```{r, echo = T}

# From assignment instructions: Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

# Find the day of the observation and assign weekday variable 1 if weekday, 2 if weekend.
for (i in 1:dim(data.imputed)[1]){
if (is.element(data.imputed$day[i], c("Saturday", "Sunday"))) {
  data.imputed$weekday[i] <- 2 # weekend
} else {
  data.imputed$weekday[i] <- 1 # weekday
}
}

# Factor weekday variable
data.imputed$weekday <- factor(data.imputed$weekday, labels = c("Weekday", "Weekend"))


# Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was creating using simulated data:

require(lattice)

# Melt data with NAs calculated. 
data.imputed.molten <- melt.data.frame(data.imputed, id.vars = c("date", "interval", "day", "weekday"), measure.vars = "steps", variable_name ="steps")

# Cast data to find average number of steps taken, averaged across all weekday days or weekend days
data.imputed.mean <- cast(data.imputed.molten, interval + weekday ~ steps, mean)

# Plot average number of steps across intervals
xyplot(data.imputed.mean$steps ~ data.imputed.mean$interval | data.imputed.mean$weekday, ylab="Number of steps", xlab="Interval", type = "l", layout = c(1,2))

```

